FOODIE, Fortran Object oriented Ordinary Differential Equations integration library based on Abstract Calculus Pattern

### 2017-05-05 papers/announcement/manuscript.tex
update paper

\begin{abstract}
The (numerical) solution of Ordinary Differential Equations (ODEs) problems is of paramount relevance, ODEs system being an ubiquitous mathematical formulation of many physical phenomena (such as those involved in fluid dynamics, chemistry, biology, evolutionary-anthropology, \dots). The present paper is the first \emph{manifesto} of FOODIE, a library aimed to numerically solve ODEs problems by means of a clear, concise and efficient \emph{abstract} interface. FOODIE, meaning Fortran Object oriented Ordinary Differential Equations integration library, has manifolds aims: to provide a set to built-in numerical schemes that are accurate, robust, validated and efficient and to allow easy application of these schemes to (almost) all ODEs problems by means of an effective Abstract Calculus Pattern. The key idea is to allow the same solver-implementation to be applied to all ODEs problems thus avoiding the re-implementation of the ODEs solver for each different ODEs problem: code re-usability is consequently maximized, FOODIE being a general robust framework. Besides, the same framework also allows rapid development of new ODEs solvers due to the high abstraction level of the library itself.

@@

>>>>

\begin{abstract}
The (numerical) solution of Ordinary Differential Equations (ODEs) problems is of paramount relevance, ODEs system being an ubiquitous mathematical formulation of many physical phenomena (such as those involved in fluid dynamics, chemistry, biology, evolutionary-anthropology, \dots). The present paper is the first \emph{manifesto} of FOODIE, a library aimed to numerically solve ODEs problems by means of a clear, concise and efficient \emph{abstract} interface. FOODIE, meaning Fortran Object oriented Ordinary Differential equations Integration Environment, has manifolds aims: to provide a set of built-in numerical schemes that are accurate, robust, validated and efficient and to allow easy application of these schemes to (almost) all ODEs problems by means of an effective Abstract Calculus Pattern. The key idea is to allow the same solver-implementation to be applied to all ODEs problems thus avoiding the re-implementation of the ODEs solver for each different ODEs problem: code re-usability is consequently maximized, FOODIE being a general robust framework. Besides, the same framework also allows rapid development of new ODEs solvers due to the high abstraction level of the library itself.

@@

\input{src/appendix/solvers-details.tex}

### 2017-05-05 papers/announcement/src/API/API.tex
update paper

As aforementioned, the programming language used is the Fortran 2008 standard, that is a minor revision of the previous Fortran 2003 standard. Such a new Fortran idioms provide (among other useful features) an almost complete support for OOP, in particular for ADT concept. Fortran 2003 has introduced the \emph{abstract derived type}: it is a derived type suitable to serve as \emph{contract} for concrete type-extensions that has not any actual implementations, rather it provides a well-defined set of type bound procedures interfaces, that in Fortran nomenclature are called \emph{deferred} procedures. Using such an abstract definition, we can implement algorithms operating on only this abstract type and on \emph{all its concrete extensions}. This is the key feature of FOODIE library: all the above described ODE solvers are implemented on the knowledge of \emph{only one abstract type}, allowing an implementation-style based on a very high-level syntax. In the meanwhile, client codes must implement their own IVPs extending only one simple abstract type.

In the subsection \ref{subsec:type_integrand} a review of the FOODIE main ADT, the \emph{integrand} type, is provided, while subsections \ref{subsec:solver_euler}, \ref{subsec:solver_tvd_rk}, \ref{subsec:solver_ls_rk}, \ref{subsec:solver_ab} and \ref{subsec:solver_leapfrog} cover the API of the currently implemented solvers.

It is worth noting that all FOODIE public \emph{entities} (ADT and solvers) must be accessed by the FOODIE module, see listing \ref{list:main_module} for an example on how access to all public FOODIE entities.

\begin{lstlisting}[firstnumber=1,style=code,caption={usage example importing all public entities of FOODIE main module},label={list:main_module}]
use foodie, only: integrand, &
                  adams_bashforth_integrator, &
                  adams_moulton_integrator, &
                  adams_bashforth_moulton_integrator, &
                  euler_explicit_integrator, &
                  leapfrog_integrator, &
                  ls_runge_kutta_integrator, &
                  tvd_runge_kutta_integrator
! or simply
use foodie
\end{lstlisting}

\subsection{The main FOODIE Abstract Data Type: the integrand type}\label{subsec:type_integrand}

The implemented ACP is based on one main ADT, the \emph{integrand} type, the definition of which is shown in listing~\ref{list:type_integrand}.

\begin{lstlisting}[firstnumber=1,style=code,caption={integrand type definition},label={list:type_integrand}]
type, abstract :: integrand
  !< Abstract type for building FOODIE ODE integrators.
  contains
    ! public deferred procedures that concrete integrand-field must implement
    procedure(time_derivative),       pass(self), deferred, public:: t
    ! operators
    procedure(symmetric_operator),   pass(lhs), deferred, public:: integrand_multiply_integrand
    procedure(integrand_op_real),    pass(lhs), deferred, public:: integrand_multiply_real
    procedure(real_op_integrand),    pass(rhs), deferred, public:: real_multiply_integrand
    procedure(symmetric_operator),   pass(lhs), deferred, public:: add
    procedure(symmetric_operator),   pass(lhs), deferred, public:: sub
    procedure(assignment_integrand), pass(lhs), deferred, public:: assign_integrand
    ! operators overloading
    generic, public:: operator(+) => add
    generic, public:: operator(-) => sub
    generic, public:: operator(*) => integrand_multiply_integrand, &
                                     real_multiply_integrand, &
                                     integrand_multiply_real
    generic, public:: assignment(=) => assign_integrand
endtype integrand
\end{lstlisting}

The \emph{integrand} type does not implement any actual integrand field, it being and abstract type. It only specifies which deferred procedures are necessary for implementing an actual concrete integrand type that can use a FOODIE solver.

As shown in listing \ref{list:type_integrand}, the number of the deferred type bound procedures that clients must implement into their own concrete extension of the \emph{integrand} ADT is very limited: essentially, there are 1 ODE-specific procedure plus some operators definition constituted by symmetric operators between 2 integrand objects, asymmetric operators between integrand and real numbers (and viceversa) and an assignment statement for the creation of new integrand objects. These procedures are analyzed in the following paragraphs.

@@

\begin{lstlisting}[firstnumber=1,style=code,caption={time derivative procedure interface},label={list:time_derivative}]
function time_derivative(self, t) result(dState_dt)
import :: integrand, R_P, I_P
class(integrand),       intent(IN) :: self      !< Integrand field.
real(R_P),    optional, intent(IN) :: t         !< Time.
class(integrand), allocatable      :: dState_dt !< Result of the time derivative function of integrand field.
endfunction time_derivative
\end{lstlisting}

This procedure-function takes two arguments, the first passed as a \emph{type bounded} argument, while the latter is optional, and it returns an integrand object. The passed dummy argument, \emph{self}, is a polymorphic argument that could be any extensions of the \emph{integrand} ADT. The optional argument \emph{t} is the \emph{time} at which the residuals function must be computed: it can be omitted in the case the residuals function does not depend directly on time.

Commonly, into the concrete implementation of this deferred abstract procedure clients embed the actual ODE equations being solved. As an example, for the Burgers equation, that is a Partial Differential Equations (PDE) system involving also a boundary value problem, this procedure embeds the spatial operator that convert the PDE to a system of algebraic ODE. As a consequence, the eventual concrete implementation of this procedure can be very complex and errors-prone. Nevertheless, the FOODIE solvers are implemented only on the above abstract interface, thus emancipating the solvers implementation from any concrete complexity.

{\color{red} Add citations to Burgers, Adams-Bashfort, leapfrog references.}

@@

\begin{lstlisting}[firstnumber=1,style=code,caption={symmetric operator procedure interface},label={list:symmetric_operator}]
function symmetric_operator(lhs, rhs) result(operator_result)
import :: integrand
class(integrand), intent(IN)  :: lhs             !< Left hand side.
class(integrand), intent(IN)  :: rhs             !< Right hand side.
class(integrand), allocatable :: operator_result !< Operator result.
endfunction symmetric_operator
\end{lstlisting}

This interface defines a class of procedures operating on 2 \emph{integrand} objects, namely it is used for the definition of the operators \emph{multiplication}, \emph{summation} and \emph{subtraction} of integrand objects. These operators are used into the above described ODE solvers, for example see equations \ref{eq:solver-euler}, \ref{eq:RK}, \ref{eq:AB} or \ref{eq:leapfrog}. The implementation details of such a procedures class are strictly dependent on the concrete extension of the integrand type. From the FOODIE solvers point of view, we need to known only that first argument passed as bounded one, the left-hand-side of the operator, and the second argument, the right-hand-side of the operator, are two integrand object and the returned object is still an integrand one.

@@

\begin{lstlisting}[firstnumber=1,style=code,caption={Integrand/real and real/integrand operators procedure interfaces},label={list:integrand_real_ops}]
function integrand_op_real(lhs, rhs) result(operator_result)
import :: integrand, R_P
class(integrand), intent(IN)  :: lhs              !< Left hand side.
real(R_P),        intent(IN)  :: rhs              !< Right hand side.
class(integrand), allocatable :: operator_result  !< Operator result.
endfunction integrand_op_real

function real_op_integrand(lhs, rhs) result(operator_result)
import :: integrand, R_P
real(R_P),        intent(IN)  :: lhs             !< Left hand side.
class(integrand), intent(IN)  :: rhs             !< Right hand side.
class(integrand), allocatable :: operator_result !< Operator result.
endfunction real_op_integrand
\end{lstlisting}

These two interfaces are necessary in order to complete the \emph{algebra} operating on the integrand object class, allowing the multiplication of an integrand object for a real number, circumstance that happens in all solvers, see equations \ref{eq:solver-euler}, \ref{eq:RK}, \ref{eq:AB} or \ref{eq:leapfrog}. The implementation details of these procedures are strictly dependent on the concrete extension of the integrand type. From the FOODIE solvers point of view, we need to known only that first argument passed as bounded one, the left-hand-side of the operator, and the second argument, the right-hand-side of the operator, are an integrand object and real number of viceversa and the returned object is still an integrand one.

@@

The abstract interface of \emph{integrand assignment} procedure is shown in listing \ref{list:integrand_assignment}.

\begin{lstlisting}[firstnumber=1,style=code,caption={integrand assignment procedure interface},label={list:integrand_assignment}]
subroutine assignment_integrand(lhs, rhs)
import :: integrand
class(integrand), intent(INOUT) :: lhs !< Left hand side.
class(integrand), intent(IN)    :: rhs !< Right hand side.
endsubroutine assignment_integrand
\end{lstlisting}

The assignment statement is necessary in order to complete the \emph{algebra} operating on the integrand object class, allowing the assignment of an integrand object by another one, circumstance that happens in all solvers, see equations \ref{eq:solver-euler}, \ref{eq:RK}, \ref{eq:AB} or \ref{eq:leapfrog}. The implementation details of this assignment is strictly dependent on the concrete extension of the integrand type. From the FOODIE solvers point of view, we need to known only that first argument passed as bounded one, the left-hand-side of the assignment, and the second argument, the right-hand-side of the assignment, are two integrand objects.

@@

\subsection{The explicit forward Euler solver}\label{subsec:solver_euler}

@@

\subsection{The explicit TVD/SSP Runge-Kutta class of solvers}\label{subsec:solver_tvd_rk}

@@

\subsection{The explicit low storage Runge-Kutta class of solvers}\label{subsec:solver_ls_rk}

@@

\subsection{The explicit Adams-Bashforth class of solvers}\label{subsec:solver_ab}

\subsection{The implicit Adams-Moulton class of solvers}\label{subsec:solver_am}

@@

\subsection{The predictor-corrector Adams-Bashforth-Moulton class of solvers}\label{subsec:solver_abm}

@@

\subsection{The leapfrog solver}\label{subsec:solver_leapfrog}

>>>>

As aforementioned, the programming language used is the Fortran 2008 (or newer) standard, that is a minor revision of the previous Fortran 2003 standard. Such a new Fortran idioms provide (among other useful features) an almost complete support for OOP, in particular for Abstract Data Type (ADT) concept. Fortran 2003 has introduced the \emph{abstract derived type}: it is a derived type suitable to serve as \emph{contract} for concrete type-extensions that has not any actual implementations, rather it provides a well-defined set of methods (in Fortran jargon they are \emph{type bound procedures}) interfaces, that in Fortran nomenclature are called \emph{deferred} procedures. Using such an abstract definition, we can implement algorithms operating on only this abstract type \emph{and on all its concrete extensions}. This is the key feature of FOODIE library: all the above described ODEs solvers are implemented on the knowledge of \emph{only one abstract type}, allowing an implementation-style based on a very high-level syntax. In the meanwhile, client codes must implement their own IVPs extending only one simple abstract type with minimal requirements.

In the subsection \ref{subsec:type_integrand} a review of the FOODIE main ADT, the \emph{integrand\_object} type, is provided, while subsections \ref{subsec:solver_euler}, \ref{subsec:solver_tvd_rk}, \ref{subsec:solver_ls_rk}, \ref{subsec:solver_ab} and \ref{subsec:solver_leapfrog} cover the API of the currently implemented solvers.

It is worth noting that all FOODIE public \emph{entities} (ADT and solvers) must be accessed by the FOODIE module, see listing \ref{list:main_module} for an example on how to access to all public FOODIE entities.

\begin{lstlisting}[firstnumber=1,style=code,caption={usage example importing all public entities of FOODIE main module},label={list:main_module}]
use foodie, only : integrand_object,                   &
                   integrator_object,                  &
                   integrator_adams_bashforth,         &
                   integrator_adams_bashforth_moulton, &
                   integrator_adams_moulton,           &
                   integrator_back_df,                 &
                   integrator_euler_explicit,          &
                   integrator_leapfrog,                &
                   integrator_lmm_ssp,                 &
                   integrator_lmm_ssp_vss,             &
                   integrator_ms_runge_kutta_ssp,      &
                   integrator_runge_kutta_emd,         &
                   integrator_runge_kutta_ls,          &
                   integrator_runge_kutta_lssp,        &
                   integrator_runge_kutta_ssp
! or simply
use foodie
\end{lstlisting}

\subsection{Main FOODIE Abstract Data Type: the integrand type}\label{subsec:type_integrand}

The implemented ACP is based on one main ADT, the \emph{integrand\_object} type, the definition of which is shown in listing~\ref{list:type_integrand}.

\begin{lstlisting}[firstnumber=1,style=code,caption={integrand type definition},label={list:type_integrand}]
type, abstract :: integrand_object
  !< Abstract type for building FOODIE ODE integrators.
  contains
    ! public deferred procedures that concrete integrand-field must implement
    procedure(time_derivative), pass(self), deferred :: t
    ! operators
    procedure(local_error_operator), pass(lhs), deferred :: local_error
    generic :: operator(.lterror.) => local_error ! Estimate local truncation error.
    procedure(symmetric_operator), pass(lhs), deferred :: integrand_add_integrand
    procedure(integrand_op_real),  pass(lhs), deferred :: integrand_add_real
    procedure(real_op_integrand),  pass(rhs), deferred :: real_add_integrand
    generic :: operator(+) => integrand_add_integrand, &
                              integrand_add_real,      &
                              real_add_integrand
    procedure(symmetric_operator),       pass(lhs), deferred :: integrand_multiply_integrand
    procedure(integrand_op_real),        pass(lhs), deferred :: integrand_multiply_real
    procedure(real_op_integrand),        pass(rhs), deferred :: real_multiply_integrand
    procedure(integrand_op_real_scalar), pass(lhs), deferred :: integrand_multiply_real_scalar
    procedure(real_scalar_op_integrand), pass(rhs), deferred :: real_scalar_multiply_integrand
    generic :: operator(*) => integrand_multiply_integrand,   &
                              integrand_multiply_real,        &
                              real_multiply_integrand,        &
                              integrand_multiply_real_scalar, &
                              real_scalar_multiply_integrand
    procedure(symmetric_operator), pass(lhs), deferred :: integrand_sub_integrand
    procedure(integrand_op_real),  pass(lhs), deferred :: integrand_sub_real
    procedure(real_op_integrand),  pass(rhs), deferred :: real_sub_integrand
    generic :: operator(-) => integrand_sub_integrand, &
                              integrand_sub_real,      &
                              real_sub_integrand
    procedure(assignment_integrand), pass(lhs), deferred :: assign_integrand
    procedure(assignment_real),      pass(lhs), deferred :: assign_real
    generic :: assignment(=) => assign_integrand, assign_real
endtype integrand_object
\end{lstlisting}

The \emph{integrand\_object} type does not implement any actual integrand field, it being and abstract type. It only specifies which deferred procedures are necessary for implementing an actual concrete integrand type that can use a FOODIE solver.

As shown in listing \ref{list:type_integrand}, the number of the deferred type bound procedures that clients must implement into their own concrete extension of the \emph{integrand} ADT is limited: essentially, there are 1 ODEs-specific procedure and some operators definition constituted by symmetric operators between 2 integrand objects, asymmetric operators between integrand and real numbers (and viceversa) and an assignment statement for the creation of new integrand objects. These procedures are analyzed in the following paragraphs.

It is worth noting that for keeping the computational efficiency high a compromise has been done in the TBPs API design: in order to avoid excessive overhead due to polymorphic entities allocation\footnote{Polymorphic allocation (and deallocation) can be very complex to optimize and it could easily generate memory leaks.} and to allows \emph{pure}\footnote{A TBP can be defined \emph{pure} when it do not have \emph{side-effects}. Purity is important for enabling compiler optimizations (otherwise prevented), for allowing easy parallelization and for imposing a clear and easy-to-debug style to the programmer.} operators, all the math operators are designed to return a real array rather than an allocatable class of \emph{integrand\_object}. The \emph{integrand\_object} is then \emph{restored} by the assignment operator that updates an instance of \emph{integrand\_object} by the real array resulting from the math operators. This is a clear restriction to the ADT abstraction level: it imposes a \emph{pattern} to the programmer that could be avoided, but, in the meanwhile, it greatly improves FOODIE performance retaining a reasonable level of abstraction.

@@

\begin{lstlisting}[firstnumber=1,style=code,caption={time derivative procedure interface},label={list:time_derivative}]
function time_derivative(self, t) result(dState_dt)
import :: integrand_object, R_P
class(integrand_object), intent(in) :: self         ! Integrand field.
real(R_P),    optional,  intent(in) :: t            ! Time.
real(R_P), allocatable              :: dState_dt(:) ! Result of the time derivative function of integrand field.
endfunction time_derivative
\end{lstlisting}

This procedure-function takes two arguments, the first passed as a \emph{type bounded} argument, while the latter is optional, and it returns a real array of kind \emph{R\_P} (that is a parametric constant of FOODIE, it being customizable at compile-time). The passed dummy argument, \emph{self}, is a polymorphic argument that could be any extensions of the \emph{integrand\_object} ADT. The optional argument \emph{t} is the \emph{time} at which the residuals function must be computed: it can be omitted in the case the residuals function does not depend directly on time.

Commonly, into the concrete implementation of this deferred abstract procedure clients embed the actual ODEs system being solved. As an example, for the Burgers equation, that is a Partial Differential Equations (PDE) system involving also a boundary value problem, this procedure embeds the spatial operator that convert the PDE to a system of algebraic ODE. As a consequence, the eventual concrete implementation of this procedure can be of any level of complexity. Nevertheless, the FOODIE solvers are implemented only on the above abstract interface, thus emancipating the solvers implementation from any concrete complexity.

@@

\begin{lstlisting}[firstnumber=1,style=code,caption={symmetric operator procedure interface},label={list:symmetric_operator}]
pure function symmetric_operator(lhs, rhs) result(operator_result)
import :: integrand_object, R_P
class(integrand_object), intent(in) :: lhs                ! Left hand side.
class(integrand_object), intent(in) :: rhs                ! Right hand side.
real(R_P), allocatable              :: operator_result(:) ! Operator result.
endfunction symmetric_operator
\end{lstlisting}

This interface defines a class of procedures operating on 2 \emph{integrand\_object} arguments, namely it is used for the definition of the operators \emph{multiplication}, \emph{summation} and \emph{subtraction} of integrand objects. These operators are used into the above described ODE solvers, for example see equations \ref{eq:solver-euler}, \ref{eq:RK}, \ref{eq:AB} or \ref{eq:leapfrog}. The implementation details of such a procedures class are strictly dependent on the concrete extension of the integrand type.

@@

\begin{lstlisting}[firstnumber=1,style=code,caption={Integrand/real and real/integrand operators procedure interfaces},label={list:integrand_real_ops}]
pure function integrand_op_real(lhs, rhs) result(operator_result)
import :: integrand_object, R_P
class(integrand_object), intent(in) :: lhs                ! Left hand side.
real(R_P),               intent(in) :: rhs(1:)            ! Right hand side.
real(R_P), allocatable              :: operator_result(:) ! Operator result.
endfunction integrand_op_real

pure function real_op_integrand(lhs, rhs) result(operator_result)
!< Asymmetric type operator `real.op.integrand`.
import :: integrand_object, R_P
class(integrand_object), intent(in) :: rhs                ! Right hand side.
real(R_P),               intent(in) :: lhs(1:)            ! Left hand side.
real(R_P), allocatable              :: operator_result(:) ! Operator result.
endfunction real_op_integrand

pure function integrand_op_real_scalar(lhs, rhs) result(operator_result)
!< Asymmetric type operator `integrand.op.real`.
import :: integrand_object, R_P
class(integrand_object), intent(in) :: lhs                ! Left hand side.
real(R_P),               intent(in) :: rhs                ! Right hand side.
real(R_P), allocatable              :: operator_result(:) ! Operator result.
endfunction integrand_op_real_scalar

pure function real_scalar_op_integrand(lhs, rhs) result(operator_result)
!< Asymmetric type operator `real.op.integrand`.
import :: integrand_object, R_P
real(R_P),               intent(in) :: lhs                ! Left hand side.
class(integrand_object), intent(in) :: rhs                ! Right hand side.
real(R_P), allocatable              :: operator_result(:) ! Operator result.
endfunction real_scalar_op_integrand
\end{lstlisting}

These foure interfaces are necessary in order to complete the \emph{algebra} operating on the integrand object class, allowing the multiplication of an integrand object for a real number (scalar or array), circumstance that happens in all solvers, see equations \ref{eq:solver-euler}, \ref{eq:RK}, \ref{eq:AB} or \ref{eq:leapfrog}. The implementation details of these procedures are strictly dependent on the concrete extension of the integrand type.

@@

The abstract interfaces of \emph{integrand assignment} procedure is shown in listing \ref{list:integrand_assignment}.

\begin{lstlisting}[firstnumber=1,style=code,caption={integrand assignment procedure interface},label={list:integrand_assignment}]
pure subroutine assignment_integrand(lhs, rhs)
import :: integrand_object
class(integrand_object), intent(inout) :: lhs ! Left hand side.
class(integrand_object), intent(in)    :: rhs ! Right hand side.
endsubroutine assignment_integrand

pure subroutine assignment_real(lhs, rhs)
!< Symmetric assignment integrand = integrand.
import :: integrand_object, R_P
class(integrand_object), intent(inout) :: lhs     ! Left hand side.
real(R_P),               intent(in)    :: rhs(1:) ! Right hand side.
endsubroutine assignment_real
\end{lstlisting}

The assignment statement is necessary in order to complete the \emph{algebra} operating on the integrand object class, allowing the
assignment of an integrand object by another one or the assignment of an integrand object and a real array (arising from the result
of other operators), circumstance that happens in all solvers, see equations \ref{eq:solver-euler}, \ref{eq:RK}, \ref{eq:AB} or \ref{eq:leapfrog}. The implementation details of this assignment is strictly dependent on the concrete extension of the integrand type.

@@

\subsection{Explicit forward Euler solver API}\label{subsec:solver_euler}

@@

\subsection{Explicit Adams-Bashforth class API}\label{subsec:solver_ab}

\subsection{Leapfrog solver}\label{subsec:solver_leapfrog}

\subsection{Explicit SSP Linear Mutistep Methods class API}

{\color{red}To be written.}

\subsection{Explicit SSP Linear Mutistep Methods with variable step-size class API}

{\color{red}To be written.}

\subsection{Explicit SSP Linear Mutistep Methods Runge-Kutta class API}

{\color{red}To be written.}

\subsection{Explicit SSP Runge-Kutta class API}\label{subsec:solver_tvd_rk}

@@

\subsection{Explicit low storage Runge-Kutta class API}\label{subsec:solver_ls_rk}

@@

\subsection{Implicit Adams-Moulton class API}\label{subsec:solver_am}

@@

\subsection{Predictor-corrector Adams-Bashforth-Moulton class API}\label{subsec:solver_abm}

@@

### 2017-05-05 papers/announcement/src/introduction/introduction.tex
update paper

Initial Value Problem (IVP, or Cauchy problem) constitutes a class of mathematical models of paramount relevance, it being applied to the modelling of a wide range of problems. Briefly, an IVP is an Ordinary Differential Equations system (ODE) coupled with specified initial values of the unknown state variables, the solution of which are searched at a given time after the initial time considered.

@@

FOODIE library must:

\begin{itemize}
  \item be written in modern Fortran (standard 2008 or newer);
  \item be written by means of Object Oriented Programming (OOP) paradigm;
  \item be well documented;
  \item be Tests Driven Developed (TDD);
  \item be collaboratively developed;
  \item be free.
\end{itemize}

FOODIE, meaning Fortran Object oriented Ordinary Differential Equations integration library, has been developed with the aim to satisfy the above specifications. The present paper is its first comprehensive presentation.

The Fortran (Formula Translator) programming language is the \emph{de facto} standard into computer science field: it strongly facilitates the effective and efficient translation of (even complex) mathematical and numerical models into an operative software without compromise on computations speed and accuracy. Moreover, its simple syntax is suitable for scientific researchers that are interested (and skilled) in the physical aspects of the numerical computations rather than computer technicians. Consequently, we develop FOODIE using Fortran language: FOODIE is written by research scientists for research scientists.

One key-point of the FOODIE development is the \emph{problem generalization}: the problem solved must be the IVP \ref{eq:IVP} rather than any of its actual definitions. Consequently, we must rely on a \emph{generic} implementation of the solvers. To this aim, OOP is very useful: it allows to express IVP \ref{eq:IVP} in a very concise and clear formulation that is really generic. In particular, our implementation is based on \emph{Abstract Calculus Pattern} (ACP) concept.

@@

Is is worth noting that this ACP emancipates the solvers implementations from any low-level problem-dependent details: the ODE solvers developed with this pattern are extremely concise, clear, maintainable and less errors-prone with respect a low-level (non abstract) pattern. Moreover, the FOODIE clients can use solvers being extremely robust: as a matter of facts, FOODIE solvers are expressed in a formulation very close to the mathematical one and are tested on an extremely varying family of problems. As shown in the following, such a great flexibility does not compromise the computational efficiency.

>>>>

Initial Value Problem (IVP, or Cauchy problem) constitutes a class of mathematical models of paramount relevance, it being applied to the modelling of a wide range of problems. An IVP is an Ordinary Differential Equation (ODE, or system of equations, ODEs) coupled with specified initial values of the unknown state variables, the solution of which are searched at a given time after the initial time considered.

@@

FOODIE development arises from the following specifications:

\begin{itemize}
  \item be written in modern Fortran (standard 2008 or newer);
  \item be abstract, thus it is written by means of Object Oriented Programming (OOP) paradigm;
  \item be well documented;
  \item be Tests Driven Developed (TDD);
  \item be collaboratively developed;
  \item be free.
\end{itemize}

Fortran (Formula Translator) programming language is the \emph{de facto} standard into computer science field: it strongly facilitates the effective and efficient translation of (even complex) mathematical and numerical models into an operative software without compromise on computations speed and accuracy. Moreover, its simple syntax is suitable for scientific researchers that are interested (and skilled) in the physical aspects of the numerical computations rather than in the information technology aspects. Consequently, we develop FOODIE using Fortran language: FOODIE is written by research scientists for research scientists.

One key-point of the FOODIE development is the \emph{problem abstraction}: the problem solved must be the IVP \ref{eq:IVP} rather than any of its real, concrete applications, e.g. the Lorentz system. Consequently, we must rely on a \emph{generic} implementation of the solvers. To this aim, OOP is very useful: it allows to express IVP \ref{eq:IVP} in a very concise and clear formulation that is really generic. In particular, our implementation is based on the \emph{Abstract Calculus Pattern} (ACP) concept.

@@

Is is worth noting that this ACP emancipates the solvers implementations from any low-level problem-dependent details: the ODE solvers developed with this pattern are extremely concise, clear, maintainable and less errors-prone with respect a low-level (non abstract) pattern. Moreover, the FOODIE clients can use solvers being extremely robust: as a matter of facts, FOODIE solvers are expressed in a formulation very close to the mathematical one and are tested on an extremely varying family of problems. As shown in the following, such a great flexibility does not compromise the computational efficiency: ACP introduces an overhead that is, however, very contained.
